---
title: "Project1_103"
output: pdf_document
date: "2024-07-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
```

## Import and summarize data


```{r}
genes <- read.csv('/Users/isabellabaldacci/Desktop/QBS103/Project/QBS103_GSE157103_genes.csv', sep = ',', header = TRUE)
metadata <- read.csv('/Users/isabellabaldacci/Desktop/QBS103/Project/QBS103_GSE157103_series_matrix.csv', header = TRUE)
dim(genes)
```


```{r}
## cleaning up metadata
metadata_clean <- dplyr::select(metadata, -c("X.Sample_submission_date","channel_count","status","last_update_date","type","channel_count","source_name_ch1","organism_ch1","apacheii","ferritin.ng.ml.","ddimer.mg.l_feu.","procalcitonin.ng.ml..","lactate.mmol.l.","fibrinogen","sofa"))
head(metadata_clean)

```
```{r}
#transposing the genes x samples matrix to samples x genes
rownames(genes) <- genes$X #assigning row names
head(genes)
genes_t <- as.data.frame(t(dplyr::select(genes, -c('X')))) #transpose as a dataframe
head(genes_t)
genes_t$participant_id <- rownames(genes_t)
head(genes_t)#assigning participant id column for later joining
which(genes_t$participant_id == "COVID_06_.y_male_NonICU") #finding column with mismatch to metadata
genes_t$participant_id[6] <- "COVID_06_:y_male_NonICU" #reassinging value to match metadata

```

```{r}
#combining the samples x genes matrix to the metadata
all_data <- dplyr::inner_join(genes_t, metadata_clean, by = 'participant_id')
head(all_data)
dim(all_data) #should be 126 x 111 (100 genes, 11 cols of metadata for all 126 samples)
```
```{r}
all_data_long <- all_data %>% tidyr::gather(key = Gene, value = count, 1:100)
all_data_long

interesting_genes <- c('AAAS', 'AAACS', 'A2M', 'AAGAB','AAK1','ABCD1','ABCG1', 'ABHD17A', 'ABHD17A','ABI1','AAMP', 'AATK','ABAT',
                       'ABCA1','ABCB10','ABHD13','ABHD14','ABHD3')
length(interesting_genes)
#colnames(all_data[101:111])
curated_genes_long <- all_data_long %>% dplyr::filter(Gene %in% interesting_genes)

#used to look for interesting genes
ggplot(all_data_long, aes(x = Gene, y = count, color = disease_status)) +
  geom_boxplot() +
  theme(
    axis.text.x = element_text(angle = 90),
    legend.position = 'top')


```


```{r}
#histogram
mean_count <- round(mean(all_data$ABHD17A),3)
sd_count <- round(sd(all_data$ABHD17A),3)


ggplot(all_data, aes(ABHD17A)) +
  geom_histogram(bins = 30 , color = 'purple1', fill = 'slateblue1') +
  theme_minimal() +
  theme(plot.title = element_text(hjust = .4, size = 20)) +
  labs(title = 'Distribution of ABHD17A count across samples', x = 'ABHD17A Expression', y = 'Frequency', size = 20) +
  scale_fill_manual(values = c('purple1')) +
  annotate(geom = 'text', x = 30, y = 12, label = paste('N:', length(all_data$ABHD17A),' samples', '\nMean Expression: ', mean_count, '\nsd: ', sd_count))

```

```{r}

#scatterplot 
colorPalette <- c('violetred', 'grey28')

all_data$crp.mg.l. <- as.numeric(all_data$crp.mg.l.)
ggplot(all_data %>% filter(crp.mg.l. != ' unknown') %>% dplyr::arrange(crp.mg.l.), aes(x = crp.mg.l. , y =  ABHD17A , color = disease_status)) +
  geom_point() +
  theme(
    axis.text.x = element_text(angle=90),
    plot.title = element_text(hjust = .4, size = 20),
    legend.position = 'bottom'
  ) +
  labs(title = 'CRP levels vs ABHD17A Expression', x = 'CRP Level (mg/L)', y = 'ABHD17A Expression') +
  scale_color_manual(labels = c('COVID', 'NOT COVID '), values = colorPalette) 
  

```

```{r}

n_covid_pos <- length(all_data[,which(all_data$disease_status == 'disease state: COVID-19')])
n_covid_neg <- length((all_data %>% dplyr::filter(disease_status != 'disease state: COVID-19'))$participant_id)

#two categorical 
dplyr::filter(curated_genes_long, Gene == 'ABHD17A') %>% 
  ggplot(aes(x = disease_status, y = count, color = 'purple', fill = mechanical_ventilation)) +
  geom_boxplot() +
  theme_minimal() + 
  theme(legend.position = 'bottom') +
  labs(title = 'Comparing ABHD17A Expression across disease status and mechanical ventilation', x = 'Disease Status', y = 'ABHD17A Expression') +
  scale_fill_manual(labels = c('no ventilator', 'ventilator'), values = c('ghostwhite', 'mediumvioletred')) +
  scale_x_discrete(labels = c('COVID', 'NON-COVID'))

```





ggplot(curated_genes_long[1:(126*8),], aes(x = Gene, y = count, fill = disease_status)) +
  geom_boxplot() +
  scale_fill_manual(labels = c('yes', 'no'),values = c('red', 'purple')) +
  theme(
    axis.text.x = element_text(angle = 90),
    legend.position = 'top')

ggplot(curated_genes_long[1:(126*8),], aes(x = Gene, y = count, fill = mechanical_ventilation)) +
  geom_boxplot() +
  scale_fill_manual(values = c('purple', 'red')) +
  theme(
    axis.text.x = element_text(angle = 90),
    legend.position = 'top')













###silly side project but not working :(
normalize <- function(vector){
  avg <- mean(vector)
  sd <- sd(vector)
  return(round(((vector-avg)/sd), 3))
}

normalized_genes <- all_data[,1:100] %>% dplyr::mutate_each(normalize)
head(normalized_genes)

```

#install.packages("corrr")
library('corrr')
#install.packages("ggcorrplot")
library(ggcorrplot)
#install.packages("FactoMineR")
library("FactoMineR")
```

head(normalized_genes)
genes.pca <- prcomp(normalized_genes[1:10,1:10])
summary(genes.pca)





