---
title: "Project1_103"
output: pdf_document
date: "2024-07-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import and summarize data


```{r}
genes <- read.csv('/Users/isabellabaldacci/Desktop/QBS103/Project/QBS103_GSE157103_genes.csv', stringsAsFactors = TRUE)
metadata <- read.csv('/Users/isabellabaldacci/Desktop/QBS103/Project/QBS103_GSE157103_series_matrix.csv', stringsAsFactors = TRUE)
```
DF1 has 100 rows and 127 columns
  col1: genes 
  col2-127: subject ids containing the following titles:
    COVID(or NonCovid)_subID_sex_ICU(or NONICU)
DF2 has 126 rows and 25 columns
  rows are subjectIDs
  Cols are: "participant_id"                         
  [2] "geo_accession"                          
  [3] "status"                                 
  [4] "X.Sample_submission_date"               
  [5] "last_update_date"                       
  [6] "type"                                   
  [7] "channel_count"                          
  [8] "source_name_ch1"                        
  [9] "organism_ch1"                           
  [10] "disease_status"                         
  [11] "age"                                    
  [12] "sex"                                    
  [13] "icu_status"                             
  [14] "apacheii" - measure of disease severity (lower == less mortality)                   
  [15] "charlson_score" - risk of death within 1 year of hospitalization for people with specific comorbid conditions                   
  [16] "mechanical_ventilation"             
  [17] "ventilator.free_days"                   
  [18] "hospital.free_days_post_45_day_followup"
  [19] "ferritin.ng.ml."                        
  [20] "crp.mg.l." - c reactive protein, produced when inflamed                          
  [21] "ddimer.mg.l_feu." - checks for blood clotting                    
  [22] "procalcitonin.ng.ml.." - high levels means body is at risk for going into sepsis (organ failure)                 
  [23] "lactate.mmol.l."- oxygen level, low values means body is not getting enough oxygen                     
  [24] "fibrinogen" - helps to heal wounds, also involved in blood clotting                            
  [25] "sofa"- sequential organ failure assessment   

```{r}
## to do: 
### choose one gene, one continuous covariate, and two categorial covariates 
### goal: generate histogram for gene expression, scatterplot for gene expression and continous covariate, boxplot of gene expression by categorial covariate 
###
```
