---
title: "Project 1: Function to Summarize and Plot Gene Expression"
output: pdf_document
date: "2024-08-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(ggpubr)
```

# Import data
```{r}
genes <- read.csv('/Users/isabellabaldacci/Desktop/QBS103/Project/QBS103_GSE157103_genes.csv',
                  sep = ',', header = TRUE)
metadata <- read.csv('/Users/isabellabaldacci/Desktop/QBS103/Project/QBS103_GSE157103_series_matrix.csv', 
                     header = TRUE)
```

## Clean up metadata
```{r}
## cleaning up metadata 
# metadata were removed if they were not helpful or if they had too many unknowns
metadata_clean <- dplyr::select(metadata, 
                                -c("X.Sample_submission_date","channel_count",
                                   "status","last_update_date","type",
                                   "channel_count","source_name_ch1",
                                   "organism_ch1","apacheii","ferritin.ng.ml.",
                                   "ddimer.mg.l_feu.","procalcitonin.ng.ml..",
                                   "lactate.mmol.l.","fibrinogen","sofa"))

```


## Transpose Genes dataframe 

```{r}
#transposing the genes x samples matrix to samples x genes
rownames(genes) <- genes$X #assigning row names
genes_t <- as.data.frame(t(dplyr::select(genes, -c('X')))) #transpose as a dataframe
genes_t$participant_id <- rownames(genes_t) #assigning participant id column for later joining
#which(genes_t$participant_id == "COVID_06_.y_male_NonICU") #finding column with mismatch to metadata
genes_t$participant_id[6] <- "COVID_06_:y_male_NonICU" #reassinging value to match metadata
```

## Combining metadata and genes dataframe

```{r}
#combining the samples x genes matrix to the metadata
all_data <- dplyr::inner_join(genes_t, metadata_clean, by = 'participant_id')

```



# Write function for plotting 

```{r}
plots <- function(df, genes_list, cont_cov, cat_cov) {
  #filter dataframe (received error message indicating to use "all of")
  df_filtered <- df %>% dplyr::select(participant_id, 
                                      all_of(genes_list), 
                                      all_of(cont_cov), 
                                      all_of(cat_cov)) 

  #cast to long
  df_filtered_long <- df_filtered %>% tidyr::pivot_longer(cols = genes_list, 
                                                          names_to = 'Gene', 
                                                          values_to = 'Expression') 

  for (gene in genes_list){ #for gene in genes list
    
    one_gene <- df_filtered_long %>% dplyr::filter(Gene == gene) #get dataframe for one gene
    
    #get data for plotting, need mean, sd, and the 75th quantile for positioning annotation box
    mean_expression <- round(mean(one_gene$Expression), 2)
    sd_expression <- round(sd(one_gene$Expression), 2)
    x_min_box <- quantile(one_gene$Expression, 3/4)
    n <- length(one_gene$Expression)
    
    #create labels
    hist_title <- paste('Distribution of ', gene, ' expression across samples')
    hist_x <- paste(gene, ' Expression')
    
    #plot histogram using theme parameters from previous assignment
    hist <- one_gene %>% ggplot(aes(Expression)) +
      geom_histogram(bins = 30, fill = 'slateblue1') +
      theme_minimal() +
      theme(plot.title = element_text(hjust = .4)) +
      labs(title = hist_title, 
       x = hist_x, 
       y = 'Frequency') +
      scale_color_manual(values = c('slateblue1')) + 
      annotate('rect', 
           xmin = x_min_box + 5, 
           xmax = x_min_box + 25, 
           ymin = 10, 
           ymax = 14, fill = 'grey', alpha = .9)+
  annotate(geom = 'text', x = x_min_box + 15, y = 12, label = paste('N:', n,' samples', 
                                                        '\n Mean Expression:', mean_expression, 
                                                        '\nsd:', sd_expression), size = 2.5)
    #print to display the histogram
    print(hist)
    
    
    #filter the data to create the scatter plot 
    #to avoid warnings of coerced NAs and to ensure cont_cov is numeric
    scatter_filtered <- one_gene %>% filter(one_gene[[cont_cov]] != ' unknown') 
    scatter_filtered[cont_cov] <- as.numeric(scatter_filtered[[cont_cov]])
    #assign the levels for crp
    if (cont_cov == 'crp.mg.l.') {
      scatter_filtered$crp.level <- cut(scatter_filtered$crp.mg.l., 
                                        breaks = c(0, 3, 10, 100, 500000), 
                                        labels = c('normal', 'moderate', 'high', 'severe'))
    }
    
    #set color palette and titles
    colorPalette_scatter <- c('azure4', 'skyblue2', 'slateblue2', 'midnightblue')
    title_scatter <- paste(toupper(cont_cov), ' vs ', gene, ' Expression')
    x_scatter <- paste(toupper(cont_cov), '(mg/L)')
    y_scatter <- paste(gene, ' Expression')
    
    #create scatterplot with parameters from previous assignment
    scatter <- scatter_filtered %>% ggplot(aes(x = .data[[cont_cov]], 
                                               y = Expression, 
                                               color = crp.level)) +
      geom_point(na.rm = TRUE) + 
      theme(
        axis.text.x = element_text(angle=90),
        plot.title = element_text(hjust = .4),
        legend.position = c(.85,.75)) +
      labs(title = title_scatter, x = x_scatter, y = y_scatter)  +
      scale_color_manual(labels = 
                           c('Normal (<3mg/L)', 'Moderate (3-10mg/L)', 
                             'High (10 -100mg/L)', 'Severe (>100mg/L)'), 
                         values = colorPalette_scatter) 
  
    #print scatterplot
    print(scatter)
    
    
    #creat titles for box plot 
    title_box <- paste(gene, ' Expression across ', cat_cov[1], ' and ', cat_cov[2])
    x_lab_box <- cat_cov[1]
    y_lab_box  <- paste(gene, ' Expression')
    #get num_covid positive and negative for annotations
    n_covid_pos <- length(dplyr::filter(
      one_gene, disease_status == 'disease state: COVID-19')$Expression)
    n_covid_neg <- length(dplyr::filter(
      one_gene, disease_status != 'disease state: COVID-19')$Expression)
    #set the y value for the annotations as the max of expression so that it will not cover the data
    y_annot <- max(one_gene$Expression)
    
    #generate box plot using same parameters as previous assignment
    box <- one_gene %>% ggplot(aes_string(x = cat_cov[1], y = 'Expression', fill = cat_cov[2])) +
      geom_boxplot() + 
      theme_minimal() + 
      theme(legend.position = 'bottom') +
      labs(title = title_box, 
           x = x_lab_box, 
           y = y_lab_box) +
      scale_fill_manual(labels = c('no ventilator', 'ventilator'), 
                        values = c('ghostwhite', 'mediumvioletred')) +
      scale_x_discrete(labels = c('COVID', 'NON-COVID')) +
      annotate(geom = 'text', x = 1, y = y_annot + 5, 
               label = paste('Covid positive: N = ', n_covid_pos), 
               color = 'black')+
      annotate(geom = 'text', x = 2, y = y_annot + 5, 
               label = paste('Covid negative: N = ', n_covid_neg), 
               color = 'black')
    print(box)
    
    #just an idea I had for putting it all together
    #all_plots <- ggpubr::ggarrange(hist, ggarrange(scatter, box, ncol = 2, 
    #labels = c('B', 'C')), nrow = 2, labels = c('A'))
      #plotlist = list(box, scatter, hist), ncol = 3, nrow = 1, widths = c(5,5,5))
    #print(all_plots)
  } 
  }
  
```

## Observing expression patterns across three genes, one continuous covariate and two categorial covariates 

### Chosen genes: ABI1, AAMP, ABHD17A
### Continuous Covariate: Crp (mg/L)
### Categorial Covariates: Disease Status and Mechanical Ventilation

```{r}

#a few warnings about deprecated methods, did not cause 
#problems so decided to suppress for final output
suppressWarnings({ plots(all_data, c('ABI1', 'AAMP', 'ABHD17A'), 
      c('crp.mg.l.'), 
      c('disease_status', 'mechanical_ventilation')) } )
  
```



